{\fontfamily{ptm}\selectfont Nesterov Type Momentum Methods}

### 2024-06-18 Notebook LaTEX/Momentum Accelerations/main.tex
>w<,added stuff to notes

\author{
    Alto
    \thanks{
        Subject type, Some Department of Some University, Location of the University,
        Country. E-mail: \texttt{author.name@university.edu}.
    }
}

@@

% ==============================================================================
\section{Preliminaries}
    This section lists foundational results important for proof in the coming sections. 
    For this section, let the ambient space be $\RR^n$ and $\Vert \cdot\Vert$ be the 2-norm until specified in the context. 
    For a general overview of smoothness and strong convexity in the Euclidean space, see \cite[theorem 2.1.5, theorem 2.1.10]{nesterov_lectures_2018} for a full exposition of the topic. 
    \subsection{Lipschitz smoothness}
        \begin{definition}[Lipschitz Smooth]
            Let $f$ be differentiable. 
            It has Lipschitz smoothness with constant $L$ if for all $x, y$
            $$
                \Vert \nabla f(x) - \nabla f(y)\Vert 
                \le 
                L \Vert x - y\Vert. 
            $$    
        \end{definition}

@@

        \begin{definition}[AG Generic PPM Form with Strong Convexity]\label{def:agg_ppm_scvx}
            Let $f$ be convex and differentiable with Lipschitz gradient and $\mu\ge0$-strongly convex, define $l_f(x; y)$ to be a function such that for all $x, \bar x \in \RR^n$: 
            \begin{align*}
               l_f(x; \bar x) + \frac{\mu}{2}\Vert x - \bar x\Vert^2
               \le f(x; \bar x) 
               \le l_f(x; \bar x) + \frac{L}{2}\Vert x - \bar x\Vert^2, 
            \end{align*}
            then for all $x_0 \in \RR^n$, and let $y_0 = x_0$, define the following variants of PPM for function $f$. 
            \begin{align*}
                x_{t + 1} &= \argmin_{x} \left\lbrace
                l_f(x; y_t) + \frac{\mu}{2}\Vert x - y_t\Vert^2 + \frac{1}{\tilde \eta_{t + 1}} \Vert x - x_t\Vert^2
                \right\rbrace, 
                \\
                y_{t + 1} &= \argmin_{x} 
                \left\lbrace
                    l_f(x; y_t) + \frac{L}{2}\Vert x - y_t\Vert^2 + 
                    \frac{1}{2\eta_{t + 1}}\Vert x - x_{t + 1}\Vert^2
                \right\rbrace. 
            \end{align*}
            Here we assume that $\eta_{t + 1} > 0$ for all $t \in \NN$. 
        \end{definition}
        \begin{remark}
            Set $\mu = 0$, the above is the same as AG Generid Triangular Form. 
        \end{remark}

@@

            In the original writing by Ahn \cite[4.9]{ahn_understanding_2022}, they didn't list what exactly this algorithm is. Here we stated that it should be equivalent to the acceleration algorithm stated in Ryu \cite[algorithm 12.1]{ryu_large-scale_2022}. 

@@

    \subsection{Scenario II}
        With the choice of $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we can recover the seminal Nesterov accelerated gradient algorithm from 1983. 
        We will discuss how other choices of $\tilde \eta_{t +1}$ results in other varients of the Accelerated gradient such as the variants proposed by Chambolle \cite{chambolle_convergence_2015}.

>>>>

\author{
    The Alto Hivemind (My Pen name)
    \thanks{
        Subject type, Some Department of Some University, Location of the University,
        Country. E-mail: \texttt{author.name@university.edu}.
    }
}

@@

% ==============================================================================
\section{Preliminaries}
    This section lists foundational results important for proof in the coming sections. 
    For this section, let the ambient space be $\RR^n$ and $\Vert \cdot\Vert$ be the 2-norm until specified in the context. 
    For a general overview of smoothness and strong convexity in the Euclidean space, see \cite[theorem 2.1.5, theorem 2.1.10]{nesterov_lectures_2018} for a full exposition of the topic. 
    \par\noindent
    We assume the reader is well-versed in convex optimization and convex analysis. 
    \subsection{Lipschitz smoothness}
        \begin{definition}[Lipschitz Smooth]
            Let $f$ be differentiable. 
            It has Lipschitz smoothness with constant $L$ if for all $x, y$
            $$
                \Vert \nabla f(x) - \nabla f(y)\Vert 
                \le 
                L \Vert x - y\Vert. 
            $$    
        \end{definition}

@@

        \begin{definition}[AG Generic PPM Form with Strong Convexity]\label{def:agg_ppm_scvx}
            Let $f$ be convex and differentiable with Lipschitz gradient and $\mu\ge0$-strongly convex, define $l_f(x; y)$ to be a function such that for all $x, \bar x \in \RR^n$: 
            \begin{align*}
               l_f(x; \bar x) + \frac{\mu}{2}\Vert x - \bar x\Vert^2
               \le f(x; \bar x) 
               \le l_f(x; \bar x) + \frac{L}{2}\Vert x - \bar x\Vert^2, 
            \end{align*}
            then for all $x_0 \in \RR^n$, and let $y_0 = x_0$, define the following variants of PPM for function $f$. 
            \begin{align*}
                x_{t + 1} &= \argmin_{x} \left\lbrace
                l_f(x; y_t) + \frac{\mu}{2}\Vert x - y_t\Vert^2 + \frac{1}{2\tilde \eta_{t + 1}} \Vert x - x_t\Vert^2
                \right\rbrace, 
                \\
                y_{t + 1} &= \argmin_{x} 
                \left\lbrace
                    l_f(x; y_t) + \frac{L}{2}\Vert x - y_t\Vert^2 + 
                    \frac{1}{2\eta_{t + 1}}\Vert x - x_{t + 1}\Vert^2
                \right\rbrace. 
            \end{align*}
            Here we assume that $\eta_{t + 1} > 0$ for all $t \in \NN$. 
        \end{definition}
        \begin{remark}
            Set $\mu = 0$, the above is the same as AG Generid Triangular Form. 
        \end{remark}

@@

            In the original writing by Ahn \cite[equation (4.9)]{ahn_understanding_2022}, they didn't identify the algorithm in the literature. 
            Here we identify it to be equivalent to the acceleration algorithm stated in Ryu \cite[algorithm 12.1]{ryu_large-scale_2022}. 

@@

    \subsection{Scenario II}
        With the choice of $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we can recover the seminal Nesterov accelerated gradient algorithm from 1983. 
        We will discuss how other choices of $\tilde \eta_{t +1}$ results in other varients of the Accelerated gradient such as the variants proposed by Chambolle \cite{chambolle_convergence_2015}. 
        To start, we make the claim that the new stepsizes results in a different algorithm. 
        \begin{lemma}[Similar Triangular Form]
            With the choice of stepszie $\tilde \eta_{t + 1} = \eta_t + L^{-1}$ as in \ref*{def:agg_tri}, we have the following specific similar triangular form of the accelerated gradient algorithm: 
                z_{t + 1} &= y_t - L^{-1} \nabla f(y_t) 
                x_{t + 1} &= z_{t + 1} + L\eta_t (z_{t + 1} - z_t)
                (1 + L\eta_{t + 1})^{-1}
                (
                x_{t + 1} + L\eta_{t + 1}z_{t + 1}
                ). 
            It also has an equivalent momentum form 
                z_{t + 1} &= y_t - L^{-1}\nabla f(y_t)
                y_{t + 1} &= z_{t + 1} + (1 + L\eta_{t + 1})^{-1}L\eta_t (z_{t + 1} - z_t). 
            To do, we show that updates sequence $x_{t + 1} = z_{t + 1} + L\eta_t (z_{t + 1} - z_t)$ is equivalent to $x_{t + 1} = x_t + \tilde\eta_{t + 1}\nabla f(y_t)$. 
            Starting with the former, susbtitute definition of $z_{t + 1}$, $z_{t + 1} = z_t + L^{-1}\nabla f(y_t)$ expanding: 
                x_{t + 1} &= y_t - L^{-1}\nabla f(y_t) 
                + L \eta_t y_t - \eta_t \nabla f(y_t) - L\eta_t z_t
                (1 + L\eta_t)y_t - (\eta_t + L^{-1})\nabla f(y_t) - L\eta_t z_t
                &= \eta_t Lz_t + x_t -(\eta_t + L^{-1}) \nabla f(y_t)  - L\eta_t z_t
                &= x_t - (\eta_t + L^{-1})\nabla f(y_t). 
            By assumption $\tilde \eta_{t + 1} = \eta_t + L^{-1}$, we ends with $x_{t + 1} = x_t + \tilde \eta_{t + 1}\nabla f(y_t)$. 
            Reducing it to the classic momentum form starts by 
                y_{t + 1} &= (1 + L\eta_{t + 1})^{-1} (x_{t + 1} + L\eta_{t + 1}z_{t + 1})
                &= (1 + L\eta_{t + 1})^{-1} (
                    z_{t + 1} + L\eta_t (z_{t + 1} - z_t) + L\eta_{t + 1} z_{t + 1}
                (1 + L\eta_{t + 1})^{-1} (
                    (1 + L\eta_{t + 1})z_{t + 1} + L\eta_t(z_{t + 1} - z_t)
                &= z_{t + 1} + (1 + L\eta_{t + 1})^{-1}L\eta_t (z_{t + 1} - z_t), 
            it negates the $x_t$ variables, therefore we have 
                z_{t + 1} &= y_t - L^{-1}\nabla f(y_t)
                y_{t + 1} &= z_{t + 1} + (1 + L\eta_{t + 1})^{-1}L\eta_t (z_{t + 1} - z_t).
            In this remark we clarify the name "similar triangule" as given in the literatures. 
            We think it is a fitting name, becaues it has a similar triangle in it. 
            We list the following observations
            \begin{enumerate}
                \item 
                The updates for $y_{t}$ from the algorithm has 
                    y_t = (1 + L\eta_t)^{-1} x_t + L\eta_t(1 + L\eta_t)^{-1} z_t, 
                therefore, $y_t$ is a convex combinations of $x_t, z_t$, so $z_t, y_t, x_t$ are three collinear points. 
                We have the ratio $\Vert y_t - x_t\Vert/\Vert z_t - y_t\Vert = L\eta_t$. 
                \item 
                The updates for $z_{t + 1}, x_{t + 1}$ are based on $y_t, x_t$, therefore vector $z_{t + 1} - y_t$ parallels to $x_{t + 1} - x_t$. 
                \item The updates for $x_{t + 1}$ has $x_{t + 1} - z_{t + 1} = L\eta_t \left(z_{t + 1} - z_t\right)$, therefore, the three points $z_t, x_{t + 1}, z_{t + 1}$ are collinear. 
                The ratio between line segment has $\Vert x_{t + 1} - z_{t + 1}\Vert/\Vert z_{t + 1} - z_t\Vert = L\eta_t$. 
            \end{enumerate}
            By these tree observations, the triangle $z_{t}, z_{t + 1}, y_t$ similar to triangle $z_t, x_{t + 1}, x_t$. 
        The next lemma will determine one choice of the stepsize sequence $\eta_t$ that recovers the classical form of Nesterov accelerated gradient method.

        \begin{lemma}[Deriving the Nesterov Sequence]
            For all $t \in \mathbb N$
                \left\lbrace
                    (1 + L\eta_{t + 1})^{-1}L \eta_t 
                    &= \frac{a_t - 1}{a_{t + 1}} , 
                    a_{t + 1}^2 - a_t &= a_t^2. 
                \right. 
            allows \hyperref[def:agg_tri]{definition \ref*{def:agg_tri}} to converge at an optimal rate. 
            The base case is $\eta_0 = 0$. 
            Continue from \ref*{lemma:smooth_agg_lyapunov_upper_bound}, With $x_{t +1} - x_t = L\tilde\eta_{t+1}(z_{t + 1} - y_t)$ for $y_t$: 
                \Upsilon_{1, t + 1}^\text{AG} &\le 
                - \frac{1}{2}\Vert x_{t+1 } - x_t\Vert^2 + 
                \frac{\tilde\eta_{t +1}L}{2}\Vert z_{t+1} - y_t\Vert^2
                - \frac{1}{2}\Vert L \tilde \eta_{t+} (z_{t+1} - y_t)\Vert^2 + 
                \frac{\tilde\eta_{t +1}L}{2}\Vert z_{t+1} - y_t\Vert^2
                    \frac{L^2\tilde\eta_{t + 1}^2}{2}
                    \frac{\tilde \eta_{t +1}}{2}
                \Vert z_{t+1} - y_t\Vert^2
                    \tilde \eta_{t+ 1} \nabla f(y_t), x_{t+1} - z_{t+1}
                \rangle. 
            This completes the analysis for $\Upsilon_{1, t + 1}^\text{AG}$ with $\tilde\eta_{t+1}$. 
            Next we consider 
                \Upsilon_{2, t + 1}^{\text{AG}} &= 
                \langle \nabla f(y_t), z_{t+1} - y_t\rangle
                \frac{L }{2} \Vert z_{t+1} - y_t\Vert^2 + 
                \langle \nabla f(y_t), z_{t + 1} - z_t\rangle
                \frac{L }{2} \Vert z_{t+1} - y_t\Vert^2
                \langle \nabla f(y_t), L^{-1}\eta_t^{-1}(x_{t+1} - z_{t + 1})\rangle
                + \frac{L}{2}\Vert z_{t + 1} - y_t\Vert^2. 
            Here we used $x_{t + 1} = z_{t + 1} + L\eta_t (z_{t + 1} - z_t)$, which was the consequence of $\tilde\eta_{t + 1} = \eta_t + L^{-1}$, as discussed from before. 
            Hence, the upper bound of the Lyapunov function for smooth AG PPM 
            (\hyperref[thm:generic_smooth_ag_convergence]{theorem \ref*{thm:generic_smooth_ag_convergence}}) 
            has 
                & \Upsilon_{1, t + 1}^{\text{AG}} + 
                \right)\Upsilon_{2, t + 1}^{\text{AG}}
                    - \frac{L^2\tilde \eta_{t + 1}^2}{2}
                    + \frac{\tilde\eta_{t+1}L}{2}
                \Vert z_{t+1} - y_t\Vert^2
                \langle \tilde \eta_{t+1} \nabla f(y_t), x_{t+1} - z_{t + 1}\rangle
                & \qquad + 
                    \langle \nabla f(y_t), L^{-1}\eta_t^{-1}(x_{t+1} - z_{t + 1})\rangle
                    + \frac{L}{2}\Vert z_{t + 1} - y_t\Vert^2
                \right). 
            If the above quantity less than or equal to zero, it will allow for convergence of the algorithm. 
            One sufficient condition is to have the coefficients for all the cross product terms to sum up to zero, and the squared normed term to sum up to a non-positive number for all $t \in \mathbb N$. 
            This gives relations 
                - \frac{1}{2}\left(
                    L^2 \tilde\eta_{t+1}^2 - \tilde\eta_{t+1}L 
                \right) + 
                \frac{L}{2}\left(
                    \sum_{i = 1}^{t} \tilde \eta_i
                \right) &\le 0
                - \left(
                    L\tilde \eta_{t + 1}^2 
                    \tilde\eta_{t + 1}
                \right) + 
                &\le 0
                -L \tilde \eta_{t+1} + 
                \sum_{i = 1}^{t + 1} \tilde \eta_i 
                &\le 0
                \sum_{i = 1}^{t + 1}\tilde \eta_i 
                &\le L \tilde \eta_{t + 1}^2, 
            and equations 
                -\tilde \eta_{t + 1} + 
                \right)L^{-1}\eta_t^{-1} 
                &= 0
                L \eta_t \tilde \eta_{t + 1}. 
            Susbtituting the equality into the inequalites we have 
                \tilde \eta_{t + 1} + \sum_{i = 1}^{t} \tilde \eta_i 
                &\le L \tilde \eta_{t + 1}^2
                \tilde \eta_{t + 1} + L\eta_t \tilde \eta_{t + 1}
                &\le L \tilde \eta_{t + 1}^2
                1 + L\eta_t &\le L \tilde\eta_{t + 1}. 
            The choice $\tilde \eta_{t + 1} = \eta_t + L^{-1}$ makes the inequality is an equality. 
            Finally, to recover the momentum stepsizes for the classical Nesterov accelerated gradient, we consider relations 
                L \sum_{i = 1}^{t + 1} \tilde \eta_i 
                &= L \tilde \eta_{t + 1} + L \sum_{i = 1}^{t} \tilde \eta_i 
                L \tilde \eta_{t + 1} + L (L \eta_t \tilde \eta_{t + 1}) 
                &= L \tilde \eta_{t + 1} + L \eta_t (L \eta_t + 1)
                &= L (\eta_t + L^{-1}) + L\eta_t (L \eta_t + 1)
                &= (L\eta_t + 1)^2, 
            Simultanenously 
                L \sum_{i = 1}^{t + 1} \tilde \eta_i 
                &= L^2 \eta_{t + 1} \tilde \eta_{t + 1}
                &= L \eta_{t + 1}(1+ L \eta_{t + 1}), 
            Combining yeilds that 
                (L\eta_t + 1)^2 
                &= L\eta_{t + 1}(1 + L \eta_{t + 1})
                &= L\eta_{t + 1} + L^2 \eta_{t + 1}^2
                (L\eta_t + 1)^2 + 1/4 &= 
                1/4 + 2(1/2)L \eta_{t + 1} + (L \eta_{t + 1})^2
                (L\eta_t + 1)^2 + 1/4 &= 
                (L \eta_{t + 1} + 1/2)^2. 
            With $a_t = L\eta_t + 1 = L \tilde \eta_t$ this is 
                a_t^2 + 1/4 &= (a_{t + 1} - 1/2)^2 
                a_t^2 + 1/4 &= a_{t + 1}^2 + 1/4 - a_{t + 1}
                a_t^2 + a_{t + 1} &= a_{t + 1}^2. 
                a_{t + 1} &= 
                \frac{1 + \sqrt{1 + 4a_t^2}}{2}
            And that is the updates for the Nesterov momentum, recall that the momentum term is linked to the PPM sequence via relation 
                (1 + L \eta_{t+ 1})^{-1}L\eta_t = L\eta_t / a_{t + 1} = (a_t - 1)/a_{t + 1}. 
            We had recovered the classic Nesterov accelerated sequence method for the smooth gradient descent algorithm. 
            This variant is distinct from the variants proposed by Chambolled \cite{chambolle_convergence_2015}, but both of which fits into the same frameworks. 
            At the time of writing, the writer is still investigating the relations between these variants of algorithms.. 
